Modeling the Linux page cache for accurate simulation of data-intensive applications

### 2020-12-10 paper_io_sim.tex

        The Linux Page Cache plays an important role in
        reducing filesystem data transfer times. With the page cache, previously
        read data can be re-read directly from memory and written data can be written to
        memory before being asynchronously flushed to disk, resulting in improved I/O performance
        on slower storage devices. The magnitude of the performance benefits that can be obtained from using the page
        cache are determined by many factors including the total amount of memory,
        the amount of data being written (i.e., dirty data), and the percentage of available memory available for
        written data. All these factors are important when determining the impact of I/O on
        application performance, particularly in data-intensive applications.

        % Swapped the order the two paragraphs and merged them into one
        The number of data-intensive applications has been steadily rising as a result of
        open-data and data sharing initiatives. Due to the sheer size of the data being
        processed, these applications must be executed on large scale infrastructure
    such as High Performance Computing (HPC) clusters or the cloud.  It
    is thus crucial to quantify the performance of these applications
    on these platforms. The goals include determining which type of hardware/software
    stacks are best suited to different application classes, as well as
    understanding the shortcomings of current algorithms, designs and
    technologies so as to propose new solutions. Unfortunately, application performance
        studies via experiments on real-world compute platforms
        faces several difficulties (high operational costs, labor-intensive experimental setups,
        shared platforms with dynamic loads that hinder reproductibilty of results) and shortcomings
        (experiments are limited to the available platform/softare configurations, which precludes
        the exploration of hypothetical scenarios). 
    Simulations address these concerns by providing simulation models
    and abstractions for the performance of computer hardware, such as
    CPU, network and storage. Consequently, simulations provide a
    cost-effective, fast, easy and reproducible way to evaluate
    application performance on arbitrary platform configurations. It thus comes
        as no surprise that, considering here only the field of parallel and distributed
        computing, a large number of simulation frameworks have been developed and used
        to support research and development endeavors~\cite{LAUNDRYLISTOFREFS_THAT_HENRI_WILL_PUT_IN}.

        Page caching is a well-known and ubiquitous technique for mitigating the I/O bottleneck
        faced by data-intensive applications. As such, it is necessary to simulate page caching
        when performing performance studies in simulation.
        Furthermore, simulating the page cache makes it possible to study relevant application-I/O
        bottlenecks that can arise over the course of execution, such as page cache
        starvation~\cite{zhuang2017}.
        While existing simulation frameworks of parallel and distributed computing
        systems  capture many relevant features of hardware/software stacks of interest, 
        they lack the ability to simulate page caches adequately. Those frameworks
        that implement some page cache simulation
        fail to capture key features such
        as dirty data and cache eviction policies~\cite{nunez2012simcan,nunez2012icancloud}. 
        Some simulators, such as that in~\cite{xu2018saving}, do capture such features,
        but are domain-specific.

@@

    In this work, we present WRENCH-cache, a simulation model of I/O
    with page cache incorporated into the WRENCH simulation framework~\cite{casanova2018wrench},
    which is  based on the popular SimGrid distributed simulation framework~\cite{XXX}.
    The added value of WRENCH-cache is to make it possible to simulate 
        application execution accurately in the presence of page caching. Our specific 
        contributions are as follows:
    \begin{compactitem}
        \item A page cache simulation model that supports 
    both single-threaded and multithreaded applications, and of both
    writeback and writethrough caches for local or network-based
    filesystems;
        \item An implementation of this model as part of the WRENCH framework; 
        \item An evaluation and validation of the accuracy of our model, and of its implementation,
              for a range of applications, platforms, and page cache configurations. 
    \end{compactitem}

>>>>

        The Linux Page Cache plays an important role in
        reducing filesystem data transfer times. With the page cache, previously
        read data can be re-read directly from memory and written data can be written to
        memory before being asynchronously flushed to disk, resulting in improved I/O performance
        on slower storage devices. The performance improvements
        depend on many factors including the total amount of memory,
        the amount of data being written (i.e., dirty data), and the amount of memory available for
        written data. All these factors are important when determining the impact of I/O on
        application performance, particularly in data-intensive applications.

        % Swapped the order the two paragraphs and merged them into one
        The number of data-intensive applications has been steadily rising as a result of
        open-data and data sharing initiatives. Due to the sheer size of the data being
        processed, these applications must be executed on large scale infrastructure
    such as High Performance Computing (HPC) clusters or the cloud.  It
    is thus crucial to quantify the performance of these applications
    on these platforms. The goals include determining which type of hardware/software
    stacks are best suited to different application classes, as well as
    understanding the shortcomings of current algorithms, designs and
    technologies so as to propose new solutions. Unfortunately, application performance
        studies via experiments on real-world compute platforms
        faces several difficulties (high operational costs, labor-intensive experimental setups,
        shared platforms with dynamic loads that hinder reproductibilty of results) and shortcomings
        (experiments are limited to the available platform/softare configurations, which precludes
        the exploration of hypothetical scenarios). 
    Simulations address these concerns by providing simulation models
    and abstractions for the performance of computer hardware, such as
    CPU, network and storage. Consequently, simulations provide a
    cost-effective, fast, easy and reproducible way to evaluate
    application performance on arbitrary platform configurations. It thus comes
        as no surprise that, considering here only the field of parallel and distributed
        computing, a large number of simulation frameworks have been developed and used
        to support research and development endeavors~\cite{ optorsim, gridsim, groudsim, cloudsim, nunez2012simcan,nunez2012icancloud, dissect_cf, cloudnetsimplusplus, fognetsimplusplus, casanova2014simgrid, ROSS, casanova2020fgcs}.

        Page caching is a well-known and ubiquitous technique for mitigating the I/O bottleneck
        faced by data-intensive applications. As such, it is necessary to simulate page caching
        when performing performance studies in simulation.
        Furthermore, simulating the page cache makes it possible to study relevant application-I/O
        bottlenecks that can arise over the course of execution, such as page cache
        starvation~\cite{zhuang2017}.
        While existing simulation frameworks of parallel and distributed computing
        systems  capture many relevant features of hardware/software stacks, 
        they lack the ability to simulate page caches adequately. Those frameworks
        that implement some page cache simulation
        fail to capture key features such
        as dirty data and cache eviction policies~\cite{nunez2012simcan,nunez2012icancloud}. 
        Some simulators, such as that in~\cite{xu2018saving}, do capture such features
        but are domain-specific.

@@

    In this work, we present WRENCH-cache, a simulation model of I/O
    with page cache incorporated into the WRENCH simulation framework~\cite{casanova2020fgcs},
    which is  based on the popular SimGrid distributed simulation framework~\cite{casanova2014simgrid}.
    The added value of WRENCH-cache is to make it possible to simulate 
        application execution accurately in the presence of page caching. Our specific 
        contributions are as follows:
    \begin{compactitem}
        \item A page cache simulation model that supports 
    both single-threaded and multithreaded applications, and of both
    writeback and writethrough caches for local or network-based
    filesystems;
        \item An implementation of this model in WRENCH; 
        \item An evaluation of the accuracy and scalability of our model, and of its implementation,
              for a range of applications, platforms, and page cache configurations. 
    \end{compactitem}
